### 2014-06-05 ApTranscriptome_TR.Rmd
Functioning RDAVIDWebService analysis. Potential bug in 'Enrichment' scores

**May 20, 2014**

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)
library(xtable)
library(methods) # required by `knitcitations` when file is `knit` file from command-line

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
library(topGO) 
library(Rgraphviz)

@@

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.only.high))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)

@@

@@

#for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
#    s <- A22.low.Ar.int.TPM.dt.sub[i]
#    X11()
#    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
#    points(s$val, s$TPM, pch = 8, col = "red")
#    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh"#) else {
#        print("OnLow")
#    }
#}

>>>>

**June 2, 2014**

@@

# Load libraries
library(R.utils)
library(ggplot2)
library(knitr)
library(knitcitations)
library(pander)
library(stringr)
library(data.table)
library(RCurl)
library(plyr)
library(MASS)
library(xtable)
library(methods) # required by `knitcitations` when file is `knit` from command-line
library(reshape2)
library(psych)

# Load Bioconductor libraries
#source("http://bioconductor.org/biocLite.R")
#biocLite("topGO")
#biocLite("Rgraphviz")
#biocLite("RDAVIDWebService")
#biocLite("GOSemSim")
library(topGO) 
library(Rgraphviz)
library(RDAVIDWebService)
library(GOSemSim)

@@

```{r gsea_by_category}
# A22 'High' genes
A22.geneList.high <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high) <- names(Ap.geneList)
A22.geneList.high[(which(names(A22.geneList.high) %in% A22.high))] <- 1
A22.high.gsea <- gsea(genelist = A22.geneList.high, geneID2GO = geneID2GO)
```

```{r gsea_setdiff, results = 'hide'}
# A22 'High' genes not in Ar**
A22.geneList.high.only <- rep(0, length = length(Ap.geneList))
names(A22.geneList.high.only) <- names(Ap.geneList)
A22.geneList.high.only[(which(names(A22.geneList.high.only) %in% A22.only.high))] <- 1
A22.high.only.gsea <- gsea(genelist = A22.geneList.high.only, geneID2GO = geneID2GO)

@@

For many of the responsive categories, it appears that there is considerable redundancy in the enriched GO terms. I use the [GOSemSim](http://www.bioconductor.org/packages/release/bioc/html/GOSemSim.html) package to determine semantic similarity of enriched GO terms `r citep("10.1093/bioinformatics/btq064")` and then perform hierarhichal clustering based on this information distance.

```{r GOSemSim}
# similarity among enriched GO terms
A22.GO.term.sim <- termSim(A22.low.gsea$GO.ID, A22.low.gsea$GO.ID, ont = "BP", organism = "fly")
# distance matrix
dist.A22.GO.term.sim <- dist(A22.GO.term.sim)
plot(hclust(dist(A22.GO.term.sim)))

```

## DAVID Gene Functional Annotation

As an alternative to topGO, in this section I use the [DAVID Functional Annotation web resource](http://david.abcc.ncifcrf.gov/) throught the Bioconductor [RDAVIDWebService](http://www.bioconductor.org/packages/release/bioc/html/RDAVIDWebService.html) package. Installing this packages is troublesome at best and my notes on how to do so are here ....

BLAST to *Apis mellifera* and *Drosophila* RefSeq Protein database

~~~
blastx
...
~~~

First, I registered as a user on the [DAVID webservice](http://david.abcc.ncifcrf.gov/content.jsp?file=DAVID_WebService.html).

I then connect and upload my complete transcriptome BLAST table as the gene background list.

Documentation for RDAVIDWebService is available [here](http://www.bdmg.com.ar/?page_id=695)

```{r DAVID_background}
# set up connection
david.connect <- DAVIDWebService$new(email = "johnsg@uvm.edu")

# load gene list
ApTranscriptome.Amel.Dros.refseq.prot <- read.table("results/trinity-full/blast_amel_dros_refseq_protein/ApTranscriptome_parallel_blastx_refseq_protein_amellifera_drosophila1.txt", sep = "\t", header = FALSE)
ApTranscriptome.Amel.Dros.refseq.prot <- data.table(ApTranscriptome.Amel.Dros.refseq.prot)
setnames(ApTranscriptome.Amel.Dros.refseq.prot, c("Transcript", "gi", "percID", "alnLength", "mismatchCount", "gapOpenCount", "queryStart", "queryEnd", "subjectStart", "subjectEnd", "eVal", "bitScore"))
setkey(ApTranscriptome.Amel.Dros.refseq.prot, Transcript)
str(ApTranscriptome.Amel.Dros.refseq.prot)

# extract RefSeq protein ID list
split.refseq.df <- colsplit(string = ApTranscriptome.Amel.Dros.refseq.prot$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
# remove version information for each gene accession
refseq.col <- strsplit(split.refseq.df$XP, split = "\\.")
refseq.id <- unlist(refseq.col)[2*1:length(refseq.col)-1]
str(refseq.id)
#write.table(refseq.id, "DAVID.background.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)

# add background list to DAVID
david.background <- addList(david.connect, refseq.id,
                      idType = "REFSEQ_PROTEIN",
                      listName = "ApTranscriptome_Amellifera_Drosophila_refseq_protein",
                      listType = "Background")
```

Next, I need to add candidate gene lists for each response type. Extract these from the full BLAST table.

```{r DAVID_A22_high}
A22.high.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.high]
# most of these were were not BLASTed. extract those that did
A22.high.geneList.blast <- A22.high.geneList[which(!is.na(A22.high.geneList$gi)), ]
# extract RefSeq protein ID
A22.high.geneList.blast.XP <- colsplit(string = A22.high.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
# remove version information for each gene accession
A22.high.geneList.refseq.col <- strsplit(A22.high.geneList.blast.XP$XP, split = "\\.")
A22.high.geneList.refseq.id <- unlist(A22.high.geneList.refseq.col)[2*1:length(A22.high.geneList.refseq.col)-1]

# add genelist to DAVID
david.A22.high.geneList <- addList(david.connect, A22.high.geneList.refseq.id,
                      idType = "REFSEQ_PROTEIN",
                      listName = "ApTranscriptome_A22_high",
                      listType = "Gene")

# set annotation category and background gene list
#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
getBackgroundListNames(david.connect)
setCurrentBackgroundPosition(david.connect, 1)
getCurrentBackgroundListPosition(david.connect)
david.connect

# Note that the current display for "species" shows only the first in list, but I confirmed in email with the programmer, Cristobal Fresno Rodriguez, that this is a bug and all species are being used, as shown in this code

getCurrentSpeciesPosition2<-function(object){
    ans <- NA
    if(length(getSpecieNames(object)) != 0){
        stub<-getStub(object)
        ans<-as.integer(strsplit(stub$getCurrentSpecies(), split = ",")[[1]]) + 1
    }
    return(ans)
}

getCurrentSpeciesPosition2(david.connect)

# obtain Functional Annotation Cluster
A22.high.termCluster <- getClusterReport(david.connect, type = "Term")
summary(A22.high.termCluster)
members(A22.high.termCluster)[[1]]$Term
members(A22.high.termCluster)[[2]]$Term
members(A22.high.termCluster)[[3]]$Term

# calculate Enrichment
geometric_mean <- function(x) exp(mean(log(x)))
-log10(geometric_mean(members(A22.high.termCluster)[[1]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[2]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[3]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[4]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[5]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[6]]$PValue))
-log10(geometric_mean(members(A22.high.termCluster)[[7]]$PValue))
```

Three clusters identified for *A. picea* genes turned on at high temperatures.

1: protein synthesis
2: regulation
3: metabolism

A22 low

```{r DAVID_A22_low}
A22.low.geneList <- ApTranscriptome.Amel.Dros.refseq.prot[A22.low]
# most of these were were not BLASTed. extract those that did
A22.low.geneList.blast <- A22.low.geneList[which(!is.na(A22.low.geneList$gi)), ]
# extract RefSeq protein ID
A22.low.geneList.blast.XP <- colsplit(string = A22.low.geneList.blast$gi, pattern="\\|", names = c("gi", "ID", "ref", "XP"))
# remove version information for each gene accession
A22.low.geneList.refseq.col <- strsplit(A22.low.geneList.blast.XP$XP, split = "\\.")
A22.low.geneList.refseq.id <- unlist(A22.low.geneList.refseq.col)[2*1:length(A22.low.geneList.refseq.col)-1]
str(A22.low.geneList.refseq.id)

# add genelist to DAVID
david.A22.low.geneList <- addList(david.connect, A22.low.geneList.refseq.id,
                      idType = "REFSEQ_PROTEIN",
                      listName = "ApTranscriptome_A22_low",
                      listType = "Gene")
david.A22.low.geneList

# set annotation category and background gene list
#setAnnotationCategories(david.connect, "GOTERM_BP_ALL")
getBackgroundListNames(david.connect)
setCurrentBackgroundPosition(david.connect, 1)
getCurrentBackgroundListPosition(david.connect)
david.connect
getCurrentSpeciesPosition2(david.connect)

# obtain Functional Annotation Cluster
A22.low.termCluster <- getClusterReport(david.connect, type = "Term")
summary(A22.low.termCluster)
members(A22.low.termCluster)[[1]]$Term
members(A22.low.termCluster)[[2]]$Term
members(A22.low.termCluster)[[3]]$Term
members(A22.low.termCluster)[[1]]$Term
members(A22.low.termCluster)[[2]]$Term
members(A22.low.termCluster)[[3]]$Term

```

7 clusters for *A. picea* low-temperature activated genes:

1. 
2.
3. 
4. membrane
5. cytochrome P450
6. zinc finger
7. Ras/Rho protein signaling

@@

for(i in unique(A22.low.Ar.int.TPM.dt.sub$Transcript)[1:10]) {
    s <- A22.low.Ar.int.TPM.dt.sub[i]
    X11()
    plot(s$val, s$pTPM, ylim = c(0, max(s$pTPM)+.1))
    points(s$val, s$TPM, pch = 8, col = "red")
    if(s[which(s$val == 38.5) , pTPM] == min(s$pTPM)) print("OffHigh") else {
        print("OnLow")
    }
}
